Tight Generalization Bounds for Hard Parameter Sharing in High-dimensional Linear Regression

### 2020-10-08 arxiv/proof_examples.tex
comments B.2

\paragraph{Part $2$: reducing the random-effects model to Theorem \ref{thm_main_RMT}.}
Using the concentration estimates in Lemma \ref{largedeviation}, we first show that $g(x)$ and its conditional expectation over $X^{(1)}, X^{(2)}$ are close. %$\exarg{\epsilon^{(1)}, \epsilon^{(2)}}{g(x) \mid X_1, X_2 ,\beta_1,\beta_2}$.
 \begin{claim}\label{claim_largedev1}
 In the setting of Corollary \ref{cor_MTL_loss}, we have that with high probability
$$\left| g(x) - \exarg{\epsilon^{(1)}, \epsilon^{(2)} ,\beta_1,\beta_2}{g(x) \mid X_1, X_2 }\right| \le p^{1/2+c} \left(\sigma^2 + \kappa^2+d^2 \right)$$.
%for any small constant $c>0$.
	% \HZ{use consistent $e_1, e_2, e$, etc. Why not just use $\e$? Don't create too many different symbols. It's bad style..}
 \end{claim}
 \begin{proof}
By Fact \ref{lem_minv} (iii), we have that with high probability,
 \be\label{op_X12}
\|X^{(1)}\|\le \sqrt{(\sqrt{n_1} + \sqrt{p})^2 + n_1 \cdot p^{-c_{\varphi}}}\lesssim \sqrt{p}, \quad \|X^{(2)}\|\le \sqrt{(\sqrt{n_2} + \sqrt{p})^2 + n_2 \cdot p^{-c_{\varphi}}}\lesssim \sqrt{p}, 
 \ee
 and
 \be\label{op_Sig1}
\| \hat\Sigma(x)^{-1}\|\le \frac1{x^2[(\sqrt{n_1} - \sqrt{p})^2 - n_1 \cdot p^{-c_{\varphi}}]+[(\sqrt{n_2} - \sqrt{p})^2 - n_2 \cdot p^{-c_{\varphi}}] }\lesssim \frac{1}{(x^2+1)p},
 \ee
where we used that $1+\tau\le \rho_1=n_1/p\le \tau^{-1}$ and $1+\tau\le \rho_2=n_2/p\le \tau^{-1}$ for a small constant $\tau>0$.

@@

We provide the predication loss using the optimal $\hat B$ in equation \eqref{eq_Bhat_2task} and optimal $\hat x=\hat x_1/\hat x_2$:
\begin{align}
L(\hat{\beta}_2^{\MTL}) &=\left\|\Sigma_2^{1/2} \left( \hat B \hat x_2 - \beta_2\right)\right\| \nonumber\\
&=  \left\| (\Sigma^{(2)})^{1/2}\hat \Sigma(\hat x)^{-1} \left[(X^{(1)})^\top X^{(1)} (\hat x\beta_1-\hat x^2\beta_2)+ (X^{(2)})^\top \epsilon^{(2)} + \hat x   (X^{(1)})^\top \epsilon^{(1)} \right]\right\|^2.\label{Lbeta_HPS}
\end{align}
%where we denoted $$\hat \Sigma:= \hat x^2 (X^{(1)})^\top X^{(1)}  + (X^{(2)})^\top X^{(2)} .$$
To illustrate the idea of bias-variance decomposition, we assume that $\hat x$ is independent of $\epsilon^{(1)}$, $\epsilon^{(2)}$, $\beta_1$ and $\beta_2$ for now. Then we can calculate the partial expectation of $L(\hat{\beta}_2^{\MTL})$ over the randomness in $\epsilon_1$ and $\epsilon_2$:
 \begin{align}
	\exarg{\epsilon^{(1)}, \epsilon^{(2)}}{L(\hat{\beta}_2^{\MTL}) \mid X_1, X_2,\beta_1,\beta_2}  &=  \bignorm{(\Sigma^{(2)})^{1/2}\hat\Sigma(\hat x)^{-1} (X^{(1)})^{\top}X^{(1)} (\hat x \beta_1 - \hat x^2\beta_2)}^2 \nonumber\\ %\label{eq_bias_2task} 
			& + \sigma^2\cdot \bigtr{ X^{(2)} \hat\Sigma(\hat x)^{-1}\Sigma^{(2)}\hat\Sigma(\hat x)^{-1} (X^{(2)})^\top} \nonumber\\
			&+ \sigma^2\cdot \hat x^2 \bigtr{ X^{(1)} \hat\Sigma(\hat x)^{-1}\Sigma^{(2)}\hat\Sigma(\hat x)^{-1} (X^{(1)})^\top}\nonumber\\
			%+\sigma^2\cdot \bigtr{\Sigma_2(\frac{W_1^2}{W_2^2} X_1^{\top}X_1 + X_2^{\top}X_2)^{-1} } \nonumber\\ 
			&= \bignorm{(\Sigma^{(2)})^{1/2}\hat\Sigma(\hat x)^{-1} (X^{(1)})^{\top}X^{(1)} (\hat x \beta_1 - \hat x^2\beta_2)}^2 +\sigma^2  \tr\left(\Sigma^{(2)}\hat \Sigma(\hat x)^{-1}\right).\nonumber
			%\label{eq_var_2task}
\end{align}
This gives a bias-variance decomposition of the MTL prediction loss: the first term on the right-hand side is the bias term, while the second term is the variance term. For the random-effects model, we further take the partial expectation over $\beta_1$ and $\beta_2$: 
 \begin{align}
	& \exarg{\epsilon^{(1)}, \epsilon^{(2)},\beta_1,\beta_2}{L(\hat{\beta}_2^{\MTL}) \mid X_1, X_2} \nonumber\\
	&	=  \hat x^2 \left[(\hat x-1)^2 \kappa^2 + \frac{(\hat x^2+1)d^2}2\right] \cdot p^{-1}\bigtr{ (X^{(1)})^{\top}X^{(1)} \hat\Sigma(\hat x)^{-1} \Sigma^{(2)} \hat\Sigma(\hat x)^{-1} (X^{(1)})^{\top}X^{(1)} }\nonumber \\
	&+\sigma^2 \tr\left(\Sigma^{(2)}\hat \Sigma(\hat x)^{-1}\right).
			\label{eq_var_2task}
\end{align}
One can compare the above expression with the predication loss for STL, that is,
\begin{align}
	\exarg{\epsilon_2}{L(\hat{\beta}_2^{\STL}) \mid X_2} = \sigma^2 \cdot \bigtr{\Sigma^{(2)} (X_2^{\top} X_2)^{-1}}, \label{eq_var_stl}
\end{align}
we observe that while the bias of $\hat{\beta}_2^{\MTL}$ is always larger than that of $\hat{\beta}_2^{\STL}$, which is zero, the variance of $\hat{\beta}_2^{\MTL}$ is always lower than that of $\hat{\beta}_2^{\STL}$.\footnote{To see why this is true, we apply the Woodbury matrix identity over the second term on the right-hand side of equation \eqref{eq_var_2task} and use the fact that for the product of two PSD matrices, its trace is always nonnegative.}
In other words, training both tasks together helps predict the target task by reducing variance while incurring a bias.
Therefore, whether multi-task learning outperforms single-task learning is determined by the bias-variance decomposition!

>>>>

\paragraph{Part $2$: reducing the random-effects model to Theorem \ref{thm_main_RMT}.}
\HZ{restructure this part. merge the result of B.17 and B.18, and state it upfront. separate its proof into two parts}
Using the concentration estimates in Lemma \ref{largedeviation}, we first show that $g(x)$ and its conditional expectation over $X^{(1)}, X^{(2)}$ are close. %$\exarg{\epsilon^{(1)}, \epsilon^{(2)}}{g(x) \mid X_1, X_2 ,\beta_1,\beta_2}$.
 \begin{claim}\label{claim_largedev1}
 In the setting of Corollary \ref{cor_MTL_loss}, we have that with high probability
$$\left| g(x) - \exarg{\epsilon^{(1)}, \epsilon^{(2)} ,\beta_1,\beta_2}{g(x) \mid X_1, X_2 }\right| \le p^{1/2+c} \left(\sigma^2 + \kappa^2+d^2 \right)$$.
%for any small constant $c>0$.
	% \HZ{use consistent $e_1, e_2, e$, etc. Why not just use $\e$? Don't create too many different symbols. It's bad style..}
 \end{claim}
 \begin{proof}
By Fact \ref{lem_minv} (iii), we have that with high probability,
 \be\label{op_X12}
\|X^{(1)}\|\le \sqrt{(\sqrt{n_1} + \sqrt{p})^2 + n_1 \cdot p^{-c_{\varphi}}}\lesssim \sqrt{p}, \quad \|X^{(2)}\|\le \sqrt{(\sqrt{n_2} + \sqrt{p})^2 + n_2 \cdot p^{-c_{\varphi}}}\lesssim \sqrt{p}, 
 \ee
 and
 \be\label{op_Sig1}
\| \hat\Sigma(x)^{-1}\|\le \frac1{x^2[(\sqrt{n_1} - \sqrt{p})^2 - n_1 \cdot p^{-c_{\varphi}}]+[(\sqrt{n_2} - \sqrt{p})^2 - n_2 \cdot p^{-c_{\varphi}}] }\lesssim \frac{1}{(x^2+1)p},
 \ee
where we used that $1+\tau\le \rho_1=n_1/p\le \tau^{-1}$ and $1+\tau\le \rho_2=n_2/p\le \tau^{-1}$ for a small constant $\tau>0$.

@@

\HZ{just say that this calculation of B.97 is similar to B.81?}
We provide the predication loss using the optimal $\hat B$ in equation \eqref{eq_Bhat_2task} and optimal $\hat x=\hat x_1/\hat x_2$:
\begin{align}
L(\hat{\beta}_2^{\MTL}) &=\left\|\Sigma_2^{1/2} \left( \hat B \hat x_2 - \beta_2\right)\right\| \nonumber\\
&=  \left\| (\Sigma^{(2)})^{1/2}\hat \Sigma(\hat x)^{-1} \left[(X^{(1)})^\top X^{(1)} (\hat x\beta_1-\hat x^2\beta_2)+ (X^{(2)})^\top \epsilon^{(2)} + \hat x   (X^{(1)})^\top \epsilon^{(1)} \right]\right\|^2.\label{Lbeta_HPS}
\end{align}
%where we denoted $$\hat \Sigma:= \hat x^2 (X^{(1)})^\top X^{(1)}  + (X^{(2)})^\top X^{(2)} .$$
To illustrate the idea of bias-variance decomposition, we assume that $\hat x$ is independent of $\epsilon^{(1)}$, $\epsilon^{(2)}$, $\beta_1$ and $\beta_2$ for now. Then we can calculate the partial expectation of $L(\hat{\beta}_2^{\MTL})$ over the randomness in $\epsilon_1$ and $\epsilon_2$:
 \begin{align}
	\exarg{\epsilon^{(1)}, \epsilon^{(2)}}{L(\hat{\beta}_2^{\MTL}) \mid X_1, X_2,\beta_1,\beta_2}  &=  \bignorm{(\Sigma^{(2)})^{1/2}\hat\Sigma(\hat x)^{-1} (X^{(1)})^{\top}X^{(1)} (\hat x \beta_1 - \hat x^2\beta_2)}^2 \nonumber\\ %\label{eq_bias_2task} 
			& + \sigma^2\cdot \bigtr{ X^{(2)} \hat\Sigma(\hat x)^{-1}\Sigma^{(2)}\hat\Sigma(\hat x)^{-1} (X^{(2)})^\top} \nonumber\\
			&+ \sigma^2\cdot \hat x^2 \bigtr{ X^{(1)} \hat\Sigma(\hat x)^{-1}\Sigma^{(2)}\hat\Sigma(\hat x)^{-1} (X^{(1)})^\top}\nonumber\\
			%+\sigma^2\cdot \bigtr{\Sigma_2(\frac{W_1^2}{W_2^2} X_1^{\top}X_1 + X_2^{\top}X_2)^{-1} } \nonumber\\ 
			&= \bignorm{(\Sigma^{(2)})^{1/2}\hat\Sigma(\hat x)^{-1} (X^{(1)})^{\top}X^{(1)} (\hat x \beta_1 - \hat x^2\beta_2)}^2 +\sigma^2  \tr\left(\Sigma^{(2)}\hat \Sigma(\hat x)^{-1}\right).\nonumber
			%\label{eq_var_2task}
\end{align}
This gives a bias-variance decomposition of the MTL prediction loss: the first term on the right-hand side is the bias term, while the second term is the variance term. For the random-effects model, we further take the partial expectation over $\beta_1$ and $\beta_2$: 
 \begin{align}
	& \exarg{\epsilon^{(1)}, \epsilon^{(2)},\beta_1,\beta_2}{L(\hat{\beta}_2^{\MTL}) \mid X_1, X_2} \nonumber\\
	&	=  \hat x^2 \left[(\hat x-1)^2 \kappa^2 + \frac{(\hat x^2+1)d^2}2\right] \cdot p^{-1}\bigtr{ (X^{(1)})^{\top}X^{(1)} \hat\Sigma(\hat x)^{-1} \Sigma^{(2)} \hat\Sigma(\hat x)^{-1} (X^{(1)})^{\top}X^{(1)} }\nonumber \\
	&+\sigma^2 \tr\left(\Sigma^{(2)}\hat \Sigma(\hat x)^{-1}\right).
			\label{eq_var_2task}
\end{align}
One can compare the above expression with the predication loss for STL, that is,
\begin{align}
	\exarg{\epsilon_2}{L(\hat{\beta}_2^{\STL}) \mid X_2} = \sigma^2 \cdot \bigtr{\Sigma^{(2)} (X_2^{\top} X_2)^{-1}}, \label{eq_var_stl}
\end{align}
we observe that while the bias of $\hat{\beta}_2^{\MTL}$ is always larger than that of $\hat{\beta}_2^{\STL}$, which is zero, the variance of $\hat{\beta}_2^{\MTL}$ is always lower than that of $\hat{\beta}_2^{\STL}$.\footnote{To see why this is true, we apply the Woodbury matrix identity over the second term on the right-hand side of equation \eqref{eq_var_2task} and use the fact that for the product of two PSD matrices, its trace is always nonnegative.}
In other words, training both tasks together helps predict the target task by reducing variance while incurring a bias.
Therefore, whether multi-task learning outperforms single-task learning is determined by the bias-variance decomposition!
